Computing likelihoods under the SMC for a general class of ARGs.

### 2023-11-14 paper.tex
Updates from Overleaf
update intro and discussion

The underlying genealogical history that relates a sample of individuals back in time
contains all possible information this sample can provide on the history of demography, 
selection and recombination for the population of interest. 
Reconstructing these genealogies with certainty is impossible however. 
As such, in the past the data structure capturing these inheritance path 
or Ancestral Recombination Graph (ARG), has usually been considered a nuisance 
variable to be integrated out during statistical inference 
\citep{kuhner_2000, griffiths_ancestral_1996}.

% explicitly outputs an ARG
ARGweaver, generates these ARGs explicitly again using MCMC sampling, by discretizing time 
they can apply HMM-based algorithms to efficiently generate new samples from the distribution of 
possible ARGs.

% model-based approaches vs

% recent advancements: heuristics
Recently, major advancements 
relying on heuristic rather than model-based approaches
made ARG inference scalable.
\citep{rasmussen_2014, kelleher_2019, speidel_2019, schaefer_2021, wohns_2022, 
zhang_biobank-scale_2023, zhan_towards_2023}
have raised the realistic prospect of ARG-based analysis becoming a standard 
part of the population and statistical genetics toolkit
This has also meant however that
we are now using the term ARG for a varying range of approxiate structures
as generated by these inference methods \citep{wong_a-general_2023}.

Firstly, in contrast to model-based inference methods like \argweaver, 
these models only deal with recombination in an implicit sense.

For example, where ARGweaver has 
the built in assumption of neighbouring trees being separated by single Subtree Prune 
Regraft (SPR) [REF Song et al.], this is not the case for any of the heuristic data 
structures.
Lack recombination nodes in the ARG, crucial in ability to compute likelihoods 
under Hudson. \citep{wong_a-general_2023}

Secondly, the lack of an underlying model also means that
we have no way to deal with uncertainty, an issue that is 
resolved differently by each inference tool.

Besides not having a unified output format, The degree of completeness 
in which genetic inheritance from ancestors to 
descendants is documented by each of these methods varies extensively.

This means there currently is no way to connect the output of the diverse 
range of ARG inference methods back 
to a model that would allow us to perform a probabilistic exploration of the 
ARG space around the point estimate, or likelihood-based inference 
(that does not only consider individual marginal trees).

[TO ADD] mention the SMC \citep{mcvean_approximating_2004}.
SMC and relatives are useful: lot of focus however on integrating over 
the distribution of trees to get Ne(t) [ADD REF: PSMC, MSMC]
Here the idea is to use the SMC in a different way that resonates more 
with the true direction of flow of genetic/haplotype information.
Results in a concrete way to compute likelihoods for *any* inferred ARG 
by integrating out the time to recombination.
(integrating out the recombination time)
and formulate an algorithm 
that is general in the sense that it can deal with the various levels of 
completeness that characterise the fields currents attempts at reconstucting genome-wide 
genealogies.

[TO ADD] remark on fact that we are only dealing with topology and branch length information
adding likelihood of observing particular mutation configuration given an ARG is straightforward.)

@@

@@

Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs. This work will contribute significantly 
to ARG-based analysis becoming a standard part of the population 
and statistical genetics toolkit. 
In particular, we see three main applications for the work presented here. 
Firstly, the ability to associate inferred ARGs with a likelihood given a demographic 
model and an associated set of parameters will help bridge the gap between the tremendous recent 
advances in ARG inference and the current state of the art statistical inference of 
past demography. Secondly, hybrid ARG-inference methods that combine the best of both 
heuristic and model-based inference approaches are now possible.
Thirdly, using a model we can quantify the uncertainty inherently associated 
with ARG-reconstruction.

% 1/ statistical inference
Currently most/all ARG-based inference methods base their analysis on marginal 
trees \citep{hejase_2022} [TO DO: add more refs]. The theory and associated algorithm 
provided here can function as a building block of a 
demography-aware genealogy sampler that would allow us to incorporate the information 
provided by recombination events.
More concretely, given our algorithm, any inferred ARG can be used a good starting point 
for a future MCMC-sampler. This potentially implies a massive runtime reduction by 
reducing the necessary burn-in period. Evenmore, by integrating out the exact time 
to the recombination event we further restrict the size of the ARG space that needs to 
be explored. 
This algorithm will also allow us to exploit the efficiencies of the tree sequence 
data structure (see \citep{mahmoudi_bayesian_2022}) to evaluate any new proposal.
We can readily incorporate $N_e$-changes as well as multiple populations and migration.\\

% 2/ hybrid ARG-inference
Biobank scale datasets, first few generations violate coalescent assumptions. Here we use heuristic 
method. Once number of lineages has reduced sufficiently, model-based approach can be used. One 
of the advantages of formulating this backwards in time. 
Combining ARG inference methods: tsinfer for first few generations (large sample sizes), where 
the initial phase is biased relative to the Wright-Fisher model when the sample size is large 
\citep{bhaskar_distortion_2014}, or when the considered events are very recent \citep{wakeley_gene_2012}\\

% 3/ quantifying uncertainty
Both of these future directions tie in with the ability to quantify uncertainty by sampling 
from an underlying distribution. This will in particular enable us to deal with those 
parts in both time and along the genome for which we have insufficient information to confidently 
reconstruct the ARG. For example the presence of polytomies in \tsinfer represents 
uncertainty. Systematicallty break them by sampling from the coalescent rather then randomly.
Quantifying uncertainty would not necessarily always have to be associated with dealing with 
the entire subset of all compatible genealogies. Uncertainty could also be quantified and passed 
on as metadata to a parent node representing the uncertainty in the interval spanned by the 
polytomy.

% can something similar be done for stacked recombinations?

@@

% final conclusion
\section{Data availability}

>>>>

The underlying genealogical history that relates a sample of individuals back in time
contains all possible information this sample can provide on the history of demography, 
selection and recombination for the population of interest. 
Reconstructing these genealogies with certainty is impossible however. 
As such, in the past the data structure capturing these inheritance paths 
or Ancestral Recombination Graph (ARG), has usually been considered a nuisance 
variable to be integrated out during statistical inference 
\citep{kuhner_2000, griffiths_ancestral_1996}.

% first major step forwards in ARG inference
% explicitly outputs an ARG
ARGweaver, generates these ARGs explicitly again using MCMC sampling, by discretizing time 
they can apply HMM-based algorithms to efficiently generate new samples from the distribution of 
possible ARGs.
% [TO ADD] also argweaver is the first to compute likelihoods for an ARG under the SMC, musn't forget to include this.
% model-based approaches vs

% recent advancements: heuristics
Recently, major advancements relying on heuristic rather than model-based approaches
have made ARG inference scalable.
\citep{rasmussen_2014, kelleher_2019, speidel_2019, schaefer_2021, wohns_2022, 
zhang_biobank-scale_2023, zhan_towards_2023}. These methods
have raised the realistic prospect of ARG-based analysis becoming a standard 
part of the population and statistical genetics toolkit.
This has also meant however that
we are now using the term ARG for a varying range of approximate structures
as generated by these different inference methods \citep{wong_a-general_2023}. 
Besides not having a unified output format, The degree of completeness 
in which genetic inheritance from ancestors to 
descendants is documented by each of these methods varies extensively. 
[TO DO:] discuss: each of these tools will deal with or resolve uncertainty 
in a different way.
This has complicated the ability to connect the output of this diverse range of 
ARG inference methods to a model-based deescription that would allow us to perform  
probabilistic exploration of the ARG space around the point estimate, 
or likelihood-based inference (that does not just consider individual marginal trees).\\

Recombination in particular:
In contrast to model-based inference methods like \argweaver, 
these models only deal with recombination in an implicit sense.\\

[TO DO: finish this section] For example, where ARGweaver has 
the built in assumption of neighbouring trees being separated by single Subtree Prune 
Regraft (SPR) [REF Song et al.], this is not the case for any of the heuristic data 
structures.
Lack recombination nodes in the ARG, crucial in ability to compute likelihoods 
under Hudson. \citep{wong_a-general_2023}

% then onto the SMC
[TO ADD: discuss SMC] mention the SMC \citep{mcvean_approximating_2004}.
SMC and relatives are useful: lot of focus however on integrating over 
the distribution of trees to get $N_e(t)$ [ADD REF: PSMC, MSMC]
Here, the idea is to use the SMC in a different way that resonates more 
with the true direction of flow of genetic information.
Results in a concrete way to compute likelihoods for *any* inferred ARG 
by integrating out the time to recombination.
and formulate an algorithm 
that is general in the sense that it can deal with the various levels of 
completeness that characterise the field's currents attempts [TO DO: make this sound more positive] 
at reconstucting genome-wide genealogies.

@@

Note that we are only considering the topology and branch length information of any ARG. 
Extending this expression by adding the likelihood of observing a set of mutations 
given an ARG is straightforward.

@@

Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs. By formulating the SMC backwards in time 
we were able to associate a single likelihood value with each edge in the ARG. 
The compatibility of the algorithm with the \tskit library allows us to 
exploit its associated efficiencies both for the current work and its future 
applications. This work will contribute significantly 
to ARG-based analysis becoming a standard part of the population 
and statistical genetics toolkit. 
In particular, we see three main applications for the work presented here. 
Firstly, the ability to associate inferred ARGs with a likelihood given a demographic 
model and an associated set of parameters will help bridge the gap between the tremendous recent 
advances in ARG inference and the current state of the art statistical inference of 
past demography. Secondly, hybrid ARG-inference methods that combine the best of both 
heuristic and model-based inference approaches are now possible.
Thirdly, using a model we can quantify the uncertainty inherently associated 
with ARG-reconstruction.

% 1/ statistical inference
Currently most/all ARG-based inference methods base their analysis on marginal 
trees \citep{hejase_2022} [TO DO: add more refs]. The theory and associated algorithm 
provided here can function as a building block of a 
demography-aware genealogy sampler that would allow us to incorporate the information 
provided by recombination events.
More concretely, given our algorithm, any inferred ARG can be used a good starting point 
for a future MCMC-sampler. This potentially implies a massive runtime reduction by 
reducing the necessary burn-in period. Furthermore, by integrating out the exact time 
to the recombination event we further restrict the size of the ARG space that needs to 
be explored. This MCMC-sampler would again rely on the tree sequence data structure to 
efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}). 
$N_e$-changes as well as multiple populations and migration can be readily incorporated 
in the current algorithm.\\

% 2/ hybrid ARG-inference
Hybrid backwards-in-time simulations that combine Wright-Fisher dynamics in the 
very recent past and coalescent simulations for the remaining time are a tried and tested 
approach \citep{bhaskar_distortion_2014, nelson_accounting_2020} to enable large scale 
simulations while avoiding the documented biases of the coalescent relative to the 
Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. The same approach 
could be used for ARG-inference where one can rely on a heuristic inference method for 
the recent past, while using a model-based approach beyond that cutoff point.\\

% 3/ quantifying uncertainty [unfinished]
Both of these future directions tie in with the desired ability of any ARG-related method to 
quantify the uncertainty associated with

to quantify uncertainty by sampling 
from an underlying distribution.

This will in particular enable us to deal with those 
parts in both time and along the genome for which we have insufficient information to confidently 
reconstruct the ARG. The presence of polytomies in ARGs inferred by \tsinfer, for example, 
represents uncertainty. Systematicallty breaking them by sampling from the coalescent rather 
then randomly would already be a major step forward.

Quantifying uncertainty would not necessarily always have to be associated 
with sampling 
with dealing with the entire subset of all compatible genealogies. Uncertainty could also 
be quantified and passed on as metadata to a parent node representing the uncertainty 
in the interval spanned by the polytomy.
% can something similar be done for stacked recombinations?

@@

% final conclusion

\section{Data availability}
